---
title: ""
format:
  pdf:
    documentclass: extarticle
    fontsize: 11pt
    mainfont: "Noto Sans"
    pdf-engine: xelatex
    geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
    df-print: kable
    include-in-header:
      text: |
        \usepackage{wrapfig}
        \usepackage{lipsum}
        \usepackage{xcolor}
        \usepackage{afterpage}
        \usepackage{pdfpages}
        \usepackage{fdsymbol}
        \usepackage{tikz}
        \usepackage{mathtools}
        \usepackage{paracol}
---

\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}
\begin{wrapfigure}[0]{R}[0pt]{3.9cm}
  
    \includegraphics[width=0.25\textwidth]{uofc2.png}
\end{wrapfigure}


\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
            
## \texorpdfstring{\fontsize{20pt}{24pt}\selectfont\textcolor[HTML]{e30c00}{STATISTICS 321: }\textcolor[HTML]{ffcc00}{Introduction to Probability}}{STATISTICS 321: Introduction to Probability}

            


## \texorpdfstring{\fontsize{15pt}{17pt}\selectfont\textcolor[HTML]{e30c00}{\textcopyright~Scott Robison 2025, }\textcolor[HTML]{ffcc00}{all rights reserved.}}{© Scott Robison 2025, all rights reserved.}




Let's keep it simple, at least at first. Probability boils down to two factors. 



1. Knowing/numbering **how many** distinct possible outcomes there are (or could be) in total.

2. The **relative size** of any one of those individual outcomes compared to all possibilities. 

One possible definition of probability, or how "likely," then follows:

$$P(A)=\frac{n(A)}{n(S)}$$
$n(A)$ the number of elements/outcomes that describe an interest $A$

$n(S)$ the number of ALL possible elements/outcomes. $S$ is the *sample space*.

\fontsize{9pt}{11pt}\selectfont
>*note and possible problem: this definition does "assume" each element/outcome has the same "weight" as every other possible outcome. (Uniform)
>
>**Consider** the sample space, $S$, of playing the lottery could be **$S=\{$You win!, You don't win$\}$**. Although, there are two possibilities, unfortunately, the do not have the same "likelihood" of realization as each other. (Non-Uniform)

\normalsize

Example 1.1: flipping a coin, rolling a die

\newpage

Example 1.2: Consider two players (A and B) playing a win/lose game (cannot tie) game(s) against each other. If they play three games what are the chances (likelihood or probability) that player A wins **two** of the three games? (Like flipping a coin?)

\newpage

```{r setup, include=FALSE}
# Keep defaults, then wrap them with sizing
old_source <- knitr::knit_hooks$get('source')
old_output <- knitr::knit_hooks$get('output')

knitr::knit_hooks$set(
  # CODE: keep styling, just wrap with a size if provided
  source = function(x, options){
    res <- old_source(x, options)
    sz  <- options$size
    if (!knitr::is_latex_output() || is.null(sz)) return(res)
    sprintf("\\begingroup\\%s\n%s\\endgroup\n", sz, res)
  },
  # RESULTS: keep default, then shrink/expand
  output = function(x, options){
    res <- old_output(x, options)
    if (!knitr::is_latex_output()) return(res)
    fs <- options$results_fontsize; ls <- options$results_leading
    if (!is.null(fs)) {
      if (is.null(ls)) {
        num <- as.numeric(sub("pt$", "", fs))
        ls  <- if (is.finite(num)) paste0(round(num*1.25,1),"pt") else "1.2\\baselineskip"
      }
      return(sprintf("\\begingroup\\fontsize{%s}{%s}\\selectfont\n%s\\endgroup\n", fs, ls, res))
    }
    if (!is.null(options$size)) return(sprintf("\\begingroup\\%s\n%s\\endgroup\n", options$size, res))
    res
  }
)

```





```{r,echo=TRUE, results='hide', message=FALSE, warning=FALSE}
#install.packages("probs")
library("probs")

```



```{r, results_fontsize='11pt'}
S=urnsamples(x=0:1, size=3, replace = TRUE, ordered = T)
nrow(S)
S
subset(S, X1+X2+X3 == 2)
nrow(subset(S, X1+X2+X3 == 2))
nrow(subset(S, X1+X2+X3 == 2))/nrow(S)

U=probspace(S)
Prob(U, X1+X2+X3 == 2)

N=urnsamples(x=c("win","lose"), size=3, replace = TRUE, ordered = T)
subset(N, rowSums(N == "win") == 2)
Prob(probspace(N), (X1=="win") + (X2=="win") + (X3=="win") == 2)

```



That player A wins **at least two** of the three games?

$$\\[2.0in]$$

```{r, results_fontsize='11pt'}
subset(S, X1+X2+X3 >= 2)
nrow(subset(S, X1+X2+X3 >= 2))
nrow(subset(S, X1+X2+X3 >= 2))/nrow(S)
```

\newpage

What if you believe the Players are not equally ‘good’ as each other? (Chances of winning are not uniform) Say, Player A has a 70% of winning, on any particular game. 
$$\\[3in]$$
```{r, results_fontsize='9pt'}
M=iidspace(x=c("win","lose"), ntrials=3, probs = c(.7,.3))
M
Prob(M, (X1=="win") + (X2=="win") + (X3=="win") == 0)
Prob(M, (X1=="win") + (X2=="win") + (X3=="win") == 1)
Prob(M, (X1=="win") + (X2=="win") + (X3=="win") == 2)
Prob(M, (X1=="win") + (X2=="win") + (X3=="win") == 3)
```
\newpage

Example 1.3: If you roll a 6-sided die twice what is the probability that the first number rolled is larger than the second number?
$$\\[3in]$$


```{r, results_fontsize='9pt'}
S <- iidspace(1:6, ntrials = 2, probs = rep(1,6)/6)  
Prob(S, X1 > X2)# Probability first roll > second roll  
```

```{r, results_fontsize='9pt'}
S <- iidspace(c("one","two","three", 4:6), ntrials = 2, probs = c(.2,.3,.1,0.05,0.26,0.09))#weird weight die
val <- c(one = 1, two = 2, three = 3, `4` = 4, `5` = 5, `6` = 6)# map labels -> numeric values
Prob(S, val[as.character(X1)] > val[as.character(X2)])# P(first > second)
subset(S, val[as.character(X1)] > val[as.character(X2)])
```

\newpage
Example 1.4: Event Space (Venn Diagrams and Set theory notation). Let's consider three Credit Card Types [Visa (V), Master Car(MC), American Express(AE)]. Come up with ways to word/visualize having **both** a Visa and Master Card.
$$\\[1.6in]$$

Who is counted for a Visa **or** a Master Card? 
$$\\[1.6in]$$



**Don’t** have a Visa **but** may have a Master Card **or** maybe an American Express.
$$\\[1.6in]$$
**Don't** have a Visa **or** you **don't** have a Master Card 
$$\\[1.6in]$$
**Neither** a Visa, **nor** a Master Card
$$\\[1.6in]$$
\newpage

Example 1.5: Roll 2, 6-sided, dice: Event $A=$The sum is $7$? Event $B=$The Product is $12$? Event $C=$ The First toss is a $5$?
$$\\[3in]$$

$P(A \cup B)$
$$\\[0.25in]$$
$P(A^c \cup B)$
$$\\[0.25in]$$

$P(A \cap C)$
$$\\[0.25in]$$


$P(A^c \cup C)$
$$\\[0.25in]$$


$P(A\cap(B\cup C))$
$$\\[0.25in]$$
$P(A\cup B^c)^c$
$$\\[0.25in]$$

$P(A^c\cap B)^c$
$$\\[0.25in]$$


$P(B\cup(A\cap C))$
$$\\[0.25in]$$



\newpage



Example 1.6: A card is selected from a standard 52 card deck, 4 suits:$\heartsuit, \diamondsuit, \clubsuit, \spadesuit$ and 13 denominations: A, 2, 3, 4, 5, 6, 7, 8, 9, 10, J, Q, K.  Only one card is selected what is the probability that…


$$\\[3in]$$

A Jack (J) **or** a Spade $\spadesuit$?
$$\\[0.25in]$$

Not a Jack nor a Spade (neither)
$$\\[0.25in]$$


Not getting a Jack or Not getting a Spade (No Jack or No Spade)
$$\\[0.25in]$$
Let's say you now take two cards. What is the probability of two $\spadesuit$s in a row?
\newpage

