---
title: ""
format:
  pdf:
    documentclass: extarticle
    fontsize: 11pt
    mainfont: "Noto Sans"
    pdf-engine: xelatex
    geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
    md-extensions: +fenced_divs+attributes
    include-in-header:
      text: |
        \usepackage{wrapfig}
        \usepackage{lipsum}
        \usepackage{xcolor}
        \usepackage{afterpage}
        \usepackage{pdfpages}
        \usepackage{fdsymbol}
        \usepackage{tikz}
        \usepackage{mathtools}
        \usepackage{paracol}
df-print: kable
---

\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}
\begin{wrapfigure}[0]{R}[0pt]{3.9cm}
  
    \includegraphics[width=0.25\textwidth]{uofc2.png}
\end{wrapfigure}


\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
            
## \texorpdfstring{\fontsize{20pt}{24pt}\selectfont\textcolor[HTML]{e30c00}{STATISTICS 321: }\textcolor[HTML]{ffcc00}{Introduction to Probability}}{STATISTICS 321: Introduction to Probability 2}

            


## \texorpdfstring{\fontsize{15pt}{17pt}\selectfont\textcolor[HTML]{e30c00}{\textcopyright~Scott Robison 2026, }\textcolor[HTML]{ffcc00}{all rights reserved.}}{© Scott Robison 2025, all rights reserved.}

Example 2.1: Scott and Jim take turns flipping a coin. Whoever "flips" a head first wins. Is there an advantage to going first?
$$\\[3in]$$

Example 2.2: Consider a lock with four spinners. The first lists only consonant letters, then a single digit integer, a spinner with only vowels, and finally another spinner with again only consonants. How many possible combinations does this lock have?

$$\\[3in]$$
\newpage

```{r}
#| echo: true
#| message: false
#| warning: false
#| results_fontsize: '11pt'
library(probs)

# spinner alphabets
consonants <- setdiff(LETTERS, c("A","E","I","O","U"))  # 21 consonants
digits     <- as.character(0:9)                        # 10 digits
vowels     <- c("A","E","I","O","U")                   # 5 vowels

# Cartesian product of the 4 spinners (base R)
lock <- expand.grid(S1 = consonants,
                    S2 = digits,
                    S3 = vowels,
                    S4 = consonants,
                    stringsAsFactors = FALSE)

# number of combinations
ncomb <- nrow(lock)
ncomb

# (optional) make it a probs probability space (equally-likely model)
lock_space <- probspace(lock)   # adds column 'probs' that sum to 1

# (optional) make a single-code string and sample a few codes
lock$code = paste0(lock$S1, lock$S2, lock$S3, lock$S4)

#sample(lock$code, 8)

```

Example 2.3: Consider a 5 digit number from 00000 to 99999. What is the probability the number selected is at least as large as 60000, has a "7" as the 3rd digit, and is even?

$$\\[3in]$$
```{r}
#| echo: true
#| message: false
#| warning: false
#| results_fontsize: '11pt'
library(probs)

# Sample space: all 5-digit numbers as strings with leading zeros
nums <- sprintf("%05d", 0:99999)

# Build a probability space (equally likely outcomes)
space <- probspace(data.frame(num = nums))

# Conditions:
cond1 <- as.integer(space$num) >= 60000
cond2 <- substr(space$num, 3, 3) == "7"
cond3 <- substr(space$num, 5, 5) %in% c("0","2","4","6","8")

# Combine conditions
event <- cond1 & cond2 & cond3

# Probability
prob_event <- Prob(space[event, ])#“Give me all rows of space where the corresponding 
#entry in event is TRUE, and give me all columns 
#(that’s why there’s a comma but nothing after it).”
prob_event
```

Example 2.4: Probability of drawing 3 spades in a row?

$$\\[3in]$$
\newpage

Example 2.5: A company has 30 employees. 3 different people will be selected for a weekly draw to win: \$100, \$50, or \$25. If a couple both work for this company what is the probability that they will win?

$$\\[3in]$$

```{r}
#| echo: true
#| message: false
#| warning: false
#| results_fontsize: '9pt'

library(probs)

# Employees labeled 1 to 30
employees <- 1:30

# Ordered selections of 3 without replacement (since prizes are different)
space <- urnsamples(employees, size = 3, replace = FALSE, ordered = TRUE)
space = probspace(space)

# Define the couple (say employees 1 and 2)
couple <- c(1, 2)

# Event 1: both members of the couple win
event_both <- apply(space, 1, function(draw) all(couple %in% draw))

# Event 2: at least one member of the couple wins
event_any <- apply(space, 1, function(draw) any(couple %in% draw))
# Probabilities using probs
prob_both <- Prob(space[event_both, ])
prob_any  <- Prob(space[event_any, ])
prob_both  # 1/145 = 0.0068966
prob_any   # 28/145 = 0.1931034
```


Example 2.6:4 men and 4 women are to be seated in a row of 8 chairs (numbered 1 through 8). How many arrangements are possible? How many arrangements if alternating gender? Probability of at least 2 women sitting together? 

$$\\[2.9in]$$


```{r}
#| echo: true
#| message: false
#| warning: false
#| results_fontsize: '9pt'
library(probs)
men <- paste0("M", 1:4)
women <- paste0("W", 1:4)
people <- c(men, women)
# Generate outcome space using urnsamples
space <- urnsamples(people, size = 8, replace = FALSE, ordered = TRUE)
# space is now a data frame of all 8! = 40320 arrangements
alternating_event <- apply(space, 1, function(row) {
  # pattern 1: M W M W ...
  pattern1 <- all(row[c(1,3,5,7)] %in% men & row[c(2,4,6,8)] %in% women)
  # pattern 2: W M W M ...
  pattern2 <- all(row[c(1,3,5,7)] %in% women & row[c(2,4,6,8)] %in% men)
  pattern1 | pattern2
})
alternating_arrangements <- sum(alternating_event)
alternating_arrangements  # 1152
# complement: no two women together
no_two_together_event <- apply(space, 1, function(row) {
  pos_w <- which(row %in% women)
  all(diff(pos_w) >= 2)
})
prob_at_least_two <- 1 - sum(no_two_together_event) / nrow(space)
prob_at_least_two  # 13/14 = 0.9285714
```
